---
title: "Revised Method"
format: html
---

```{r setup, include = FALSE}
library(tidyverse)
library(sf)
library(vroom)
library(here)

weight.90 <- vroom("D:/Github/ORD_Water_Supply/Water_Use/outputs/data/blocks_90_weight.csv")

# Load Water Source Result
ws.df <- vroom(here("outputs/Well_Estimates_2020_Blks.csv"), show_col_types = FALSE)

# Load CWS Boundary Data and filter to state/utility boundaries
cws <- st_read(here("Method_2/data/EPA_CWS_V1.shp"))%>%
 filter(Method == "State")

# Load system information
info.df <- vroom("D:/Github/ORD_SAB_Model/Input_Data/SDWIS/Water_System_Detail_2023Q4.csv")

# Load block join files for public water systems
join.files <- list.files("D:/Github/ORD_SAB_Model/Output_Data/Archive/Block_Join", full.names = TRUE)

join.df <- vroom(join.files)%>%
  filter(PWSID_12 %in% cws$PWSID)%>%
  group_by(GEOID20)%>%
  summarise(I_Area_km = sum(I_Area_km),
            Block_Area_km = Block_Area_km[1])

# Load population
pop <- vroom("D:/data/nhgis/tables/Blocks/nhgis0307_ds248_2020_block.csv")%>%
  select(GISJOIN,GEOID,U7B001,U7G001)%>%
  mutate(GEOID20 = substr(GEOID,10,24))%>%
  select(!GEOID)


## Load wells files
files <- list.files("D:/Github/ORD_Water_Supply/Water_Use/Outputs/data/Drilled_Wells_Allocated", pattern=".csv$", full.names=TRUE)

wells.df <- vroom(files)%>%
  select(GISJOIN,HU_2020,Est_Wells_2020)

# Check for Census Blocks that have accurate wells and a public service area.

```

## Overview

### Step 1
For States where we have > 90% of the population served by CWS within authoritative service area boundaries:
- If in service are, then public. If not, then private. 

### Step 2
For all states:
- Blocks that have zero housing units in 2020, but are within a Census designated urban area, should be classified as Public

### Step 3
**We can't do this, since the water districts in many states will overwrite way too many wells.**

For all other states with < 90% service areas:

- Reclassify well layer to binary: anything with <= 90% wells will be considered public water

- Census blocks that were classified as using private water are overridden if they are within a known service area boundary.

### Step 4
If we have well drill records for a census block, they should be added to the Census block



### Determine the percent of population served covered by authoritative boundaries from each state

```{r}
pct.pop <- info.df%>%
  select(`PWS ID`, `Primacy Agency`, `Population Served Count`)%>%
  mutate(Authoritative = ifelse(`PWS ID` %in% cws$PWSID,"YES","NO"))%>%
  group_by(`Primacy Agency`)%>%
  mutate(Pop_Served_Total = sum(`Population Served Count`, na.rm = TRUE))%>%
  ungroup()%>%
  mutate(Auth_Pop = ifelse(Authoritative == "NO",0,`Population Served Count`))%>%
  group_by(`Primacy Agency`)%>%
  summarise(Auth_Pop = sum(Auth_Pop),Pop_Served_Total = Pop_Served_Total[1])%>%
  mutate(Pct_Auth = round(100*(Auth_Pop/Pop_Served_Total)))%>%
  filter(Pct_Auth > 95)

ggplot(pct.pop)+
  geom_col(aes(x = Pct_Auth, y = fct_reorder(`Primacy Agency`,Pct_Auth)))+
  coord_cartesian(xlim = c(75,100))+
  labs(title = "Population Served by Community Water Systems\nWith Authoritative Service Areas",
       x = "% of State Population Using CWS")
```

## Step 1:

Load block centroids and intersect with Census defined urban areas. Then run an if/else statement for blocks with zero housing units that intersect urban areas.

```{r urbanAreas, eval = FALSE}

zero.units <- vroom("D:/data/nhgis/tables/Blocks/nhgis0307_ds248_2020_block.csv")%>%
  select(GISJOIN,U7G001)%>%
  filter(U7G001 == 0)

cntrds <- vroom("D:/Github/ORD_Water_Supply/data/blocks_2020_all_centroids.csv")%>%
  filter(GISJOIN %in% zero.units$GISJOIN)%>%
  st_as_sf(coords = c("X","Y"), crs = st_crs(4326))

# Import Urban Areas
ua <- st_read(here("Method_2/data/Urban_Areas.shp"))%>%
  st_transform(st_crs(4326))

# Dissolve Urban Areas
agg <- ua%>%
  summarise()%>%
  st_make_valid()

# Find block centroids that intersect urban areas
tictoc::tic()
cntrd.intrsct <- cntrds%>%
  filter(st_intersects(.,agg,sparse = FALSE))
tictoc::toc()

# Save data frame of intersecting blocks with zero housing units
df.zero.urban <- cntrd.intrsct%>%
  mutate(URBAN = TRUE)%>%
  st_drop_geometry()
```

## Step 2
### Apply Reclassification from Service Areas

```{r reclass1, eval= FALSE}
reclass.1 <- ws.df%>%
  filter(State %in% pct.pop$`Primacy Agency`)%>%
  left_join(join.df, by = c("GEOID_Blk"="GEOID20"))%>%
  mutate(I_Area_km = replace_na(I_Area_km,0),
         OldClass = ifelse(Est_Wells_2020/HU_2020>=0.5,"Private","Public"),
         NewClass = ifelse(I_Area_km > 0,"Public","Private"),
         Old_New = paste0(OldClass," -> ",NewClass))

# Save spatial File
states <- c("AZ","AR","CA","KS","KY","NH","NJ","PA","RI","TN","TX","UT","WA")

sf <- data.frame()

for(s in states){
  sf.next <- st_read(paste0("D:/data/nhgis/boundaries/Blocks/2020/",s,"_block_2020.shp"))%>%
  select(GEOID20)%>%
  st_transform(st_crs(5070))%>%
    mutate(Area_Km = as.numeric(st_area(.))/1000000)%>%
  left_join(reclass.1, by = c("GEOID20"="GEOID_Blk"))%>%
    mutate(NewClass = ifelse(I_Area_km/Area_Km > 0.1,"Public","Private"),
         Old_New = paste0(OldClass," -> ",NewClass))

  sf <- rbind(sf,sf.next)

  print(paste0("Completed ",s," @ ", Sys.time()))
}


st_write(sf,here("Method_2/data/Method_2.gpkg"), layer = "Step_2", append = FALSE)

# sf <- st_read(here("Method_2/data/Method_2.gpkg"), layer = "Step_1")
reclass.2 <- sf%>%
  st_drop_geometry()%>%
  mutate(Reclass_Step = 2)
  left_join(pop, by = "GEOID20")

# Save data table
vroom_write(reclass.2, here("Method_2/data/Reclass_2.csv"), delim = ",", append = FALSE)

```



## Step 3 (Part 2 of step 2 in paper):

For every block that was not captured in steps 1 & 2, check if the block falls within an authoritative boundary. 

- Census blocks that were classified as using private water are overridden if they are within a known service area boundary.

```{r reclass3, eval=FALSE}
reclass.2 <- vroom(here("Method_2/data/Reclass_2.csv"))

pws.ids <- unique(cws$PWSID)

auth.blks <- join.df%>%
  filter(I_Area_km/Block_Area_km > 0.1)

reclass.3 <- ws.df%>%
  filter(!GEOID_Blk %in% reclass.2$GEOID20 & GEOID_Blk %in% auth.blks$GEOID20)%>%
  mutate(OldClass = Pred_Class,
         NewClass = ifelse(GEOID_Blk %in% auth.blks$GEOID20, "Public",OldClass),
         Old_New = paste0(OldClass," -> ",NewClass),
         Reclass_Step = 3)
```

## Step 4:

In step 4, we combine the reclassification files, then calculate binary estimated wells: zero for 'public' and wells = housing units for 'private'. Then, for publicly classified blocks, we add well logs to 1990 wells to gain an estimate of current wells. If current wells exceeds 50% of current housing units, the block is flipped to private.
```{r}
df.urban.zero <- vroom(here("Method_2/data/Urban_NoHousing.csv"))%>%
  left_join(pop, by = "GISJOIN")

reclass.urban <- ws.df%>%
  filter(!GEOID_Blk %in% reclass.2$GEOID20 & !GEOID_Blk %in% reclass.3$GEOID_Blk)%>%
  filter(GEOID_Blk %in% df.urban.zero$GEOID20)%>%
  mutate(OldClass = ifelse(Est_Wells_2020/HU_2020>0.9,"Private","Public"),
         NewClass = ifelse(GEOID_Blk %in% df.urban.zero$GEOID20,"Public",
                           ifelse(Est_Wells_2020/HU_2020>0.9,"Private",OldClass)),
         Old_New = paste0(OldClass," -> ",NewClass),
         Reclass = "Urban Area")

vroom_write(reclass.urban, here("Method_2/data/Reclass_1.csv"), delim = ",", append = FALSE)

# Load well join files
well.join.files <- list.files(here("data/Well Logs/Census_Blocks/"), full.names = TRUE)
wells.join <- vroom(well.join.files)%>%
  select(GISJOIN,Well_ID,Install_Date)%>%
  group_by(Well_ID)%>%
  mutate(Install_Date = max(Install_Date))%>%
  ungroup()%>%
  distinct()%>%
  group_by(GISJOIN)%>%
  summarise(nWells = n())

## Load 1990 well estimates and add them to drilled wells
wells.90 <- vroom("D:/Github/ORD_Water_Supply/Water_Use/outputs/data/Blocks_2020.csv")%>%
  mutate(HU_1990 = replace_na(HU_1990,0),
         Pct_Wells_Other_S = replace_na(Pct_Wells_Other_S,0),
         Wells_90 = round(HU_1990 * Pct_Wells_Other_S))%>%
  select(GISJOIN,Wells_90)
  
wells.add <- wells.join%>%
  left_join(wells.90)%>%
  mutate(Total_Wells = nWells + Wells_90)


# Combine Everything and recalculate Wells
no.reclass <- ws.df%>%
  filter(!GEOID_Blk %in% reclass.urban$GEOID_Blk & !GEOID_Blk %in% reclass.2$GEOID20 & !GEOID_Blk %in% reclass.3$GEOID_Blk)%>%
  mutate(OldClass = Pred_Class,
         NewClass = ifelse(wasIsland == TRUE & Pred_Class == "Private","Public",
                            ifelse(wasIsland == TRUE & Pred_Class == "Public","Private",Pred_Class)),
         Old_New = paste0(OldClass," -> ",NewClass),
         Reclass = ifelse(OldClass == NewClass,"None","Was Island"))


## Assign Reclassification Methods
colnames(reclass.2)[1] <- "GEOID_Blk"
r0 <- no.reclass%>%
  select(GEOID_Blk,GEOID_BlkGrp,HU_2020,Prob_Pub,Tree_Cat,OldClass,NewClass,Reclass)%>%
  mutate(OldClass = as.character(OldClass),
         NewClass = as.character(NewClass),
         OldClass = replace_na(OldClass,"NONE"),
         NewClass = replace_na(NewClass,"NONE"))

r1 <- reclass.urban%>%
  filter(!GEOID_Blk %in% reclass.2$GEOID_Blk)%>%
  select(GEOID_Blk,GEOID_BlkGrp,HU_2020,Prob_Pub,Tree_Cat,OldClass,NewClass,Reclass)%>%
  mutate(OldClass = "NONE",
         NewClass = as.character(NewClass),
         NewClass = replace_na(NewClass,"NONE"))

r2 <- reclass.2%>%
  mutate(OldClass = as.character(OldClass),
         NewClass = as.character(NewClass),
         OldClass = replace_na(OldClass,"NONE"),
         NewClass = replace_na(NewClass,"NONE"),
         Reclass = ifelse(OldClass == NewClass,"None","Complete CWS Boundaries"))%>%
  select(GEOID_Blk,GEOID_BlkGrp,HU_2020,Prob_Pub,Tree_Cat,OldClass,NewClass,Reclass)

r3 <- reclass.3%>%
  mutate(OldClass = as.character(OldClass),
         NewClass = as.character(NewClass),
         OldClass = replace_na(OldClass,"NONE"),
         NewClass = replace_na(NewClass,"NONE"),
         Reclass = ifelse(NewClass == "Public" & !OldClass == NewClass,"Within CWS","None"))%>%
  select(GEOID_Blk,GEOID_BlkGrp,HU_2020,Prob_Pub,Tree_Cat,OldClass,NewClass,Reclass)

r.combine <- r0%>%
  rbind(r1,r2,r3)

check <- as.data.frame(table(r.combine$GEOID_Blk))%>%
  filter(Freq > 1)

r4 <- r.combine%>%
  left_join(pop, by = c("GEOID_Blk"="GEOID20"))%>%
  left_join(wells.add)%>%
  mutate(nWells = replace_na(nWells,0),
         Wells_90 = replace_na(Wells_90,0),
         Total_Wells = replace_na(Total_Wells,0),
         HU_2020 = replace_na(HU_2020,0),
         # OldClass = replace_na(OldClass,"NONE"),
         # NewClass = replace_na(NewClass,"NONE"),
         NewClass = ifelse((Total_Wells/HU_2020)>=0.5 & HU_2020>0,"Private",NewClass),
         Reclass = ifelse(NewClass == "Private" & OldClass == "Public" & (Total_Wells/HU_2020)>=0.5 & HU_2020>0,"Drill Logs",Reclass))
```



## Create Final Dataset for Export
```{r}
fips <- tidycensus::fips_codes%>%
  mutate(Join = paste0(state_code,county_code))

df.final <- r4%>%
  mutate(Join = substr(GEOID_Blk,1,5))%>%
  left_join(fips, by = "Join")%>%
  mutate(Est_Wells = ifelse(NewClass == "Private",HU_2020,0),
         Est_Wells = ifelse(Total_Wells > 0,Total_Wells,Est_Wells),
         Est_Wells = ifelse(Est_Wells>HU_2020,HU_2020,Est_Wells),
         Est_Wells = replace_na(Est_Wells,0))%>%
  select(state,state_code,county_code,GEOID_Blk,GEOID_BlkGrp,U7B001,U7G001,Prob_Pub,Tree_Cat,OldClass,NewClass,Reclass,Wells_90,Est_Wells)%>%
  setNames(c("State","State_Fips","County_Fips","GEOID_Block","GEOID_BlockGroup","Population","Housing_Units","Probability_Public","Tree_Node","Old_Class","Final_Class","Reclass_Type","Wells_1990","Est_Wells_2020"))%>%
  mutate(Reclass_Type = ifelse(Final_Class == Old_Class,"None",Reclass_Type),
         Pct_Public = ifelse(Housing_Units > 0, round(100*((Housing_Units-Est_Wells_2020)/Housing_Units)),NA),
         Pct_Wells = ifelse(Housing_Units > 0, round(100*(Est_Wells_2020/Housing_Units)),NA),
         Pop_Wells = ifelse(Population > 0, round((Population/Housing_Units)*Est_Wells_2020),0),
         Pop_CWS = ifelse(Population > 0 & Est_Wells_2020 == 0,Population,
                          ifelse(Population > 0 & Est_Wells_2020 > 0,round(Population - Pop_Wells),0)))

vroom_write(df.final,here("outputs/Well_Estimates_2020_Blocks_10302024.csv"),delim = ",", append=FALSE)

# Create Shapefiles
shps <- list.files("D:/data/nhgis/boundaries/Blocks/2020", pattern = ".shp$", full.names = TRUE)

for(n in 1:length(shps)){
  st.sf <- st_read(shps[n])%>%
    select(GEOID20)%>%
    setNames(c("GEOID_Block","geometry"))%>%
    st_transform(4326)%>%
    st_make_valid()%>%
    left_join(df.final)
  
  st_write(st.sf,here("outputs/Census_Blocks.gpkg"),layer = st.sf$State[1], append = FALSE)
}

```



## Statistics

```{r stats}
# Reclass Counts
reclass.counts <- df.final%>%
  drop_na(Final_Class)%>%
  mutate(nRows = n())%>%
  group_by(Reclass_Type)%>%
  summarise(N = n(),
            nRows = nRows[1])%>%
  mutate(Percent = round(100*(N/nRows),1))

write.csv(reclass.counts,here("Analysis/Reclass_Counts.csv"))
```






## Check Wells vs Public Water
Here, we compare authoritative service area boundaries and well logs which we consider accurate. We can only compare in states where we have both. Those States are:

- Arizona
- Arkansas
- California
- Kansas
- Pennsylvania
- Tennessee

"AZ","AR","CA","CT","KS","KY","NH","NJ","NM","NY","NC","PA","RI","TN","TX","UT","WA"

```{r}
# Load block join files for public water systems
join.files <- data.frame(path = list.files("D:/Github/ORD_SAB_Model/Output_Data/Archive/Block_Join", full.names = TRUE),
                         file = list.files("D:/Github/ORD_SAB_Model/Output_Data/Archive/Block_Join"))%>%
  mutate(state = substr(file,1,2))%>%
  filter(state %in% c("AZ","CA","KS","PA","TN"))

cws.blks <- vroom(join.files$path)%>%
  full_join(wells.df,by = "GISJOIN")%>%
  mutate(Pct_CWS = 100*(I_Area_km/Block_Area_km),
         Has_CWS = ifelse(is.na(PWSID_12),FALSE,
                          ifelse(Pct_CWS >=10,TRUE,FALSE)))

conflicting <- cws.blks%>%
  filter(Has_CWS == TRUE & Est_Wells_2020 > 0)%>%
  mutate(Pct_Wells = 100*(Est_Wells_2020/HU_2020))

# Save a spatial file for investigation.
cntrds <- vroom("D:/Github/ORD_Water_Supply/data/blocks_2020_all_centroids.csv")

conf.sf <- cntrds%>%
  left_join(conflicting)%>%
  filter(GISJOIN %in% conflicting$GISJOIN)%>%
  st_as_sf(coords = c("X","Y"), crs = st_crs(4326))

st_write(conf.sf,"D:/temp/conflicting.gpkg",layer = "points", append = FALSE)
```


```{r reclass1Result}
reclass.1 <- vroom(here("Method_2/data/Reclass_1.csv"))

stats <- reclass.1%>%
  group_by(State, Old_New)%>%
  summarise(n = n(),
            Area = sum(Area_Km,na.rm=TRUE),
            Pop = sum(U7B001, na.rm = TRUE))%>%
  drop_na()%>%
  ungroup()

ggplot(stats)+
  geom_col(aes(x = n, y = Old_New))+
  facet_wrap(~State, scales = "free", ncol = 3)

ggplot(stats)+
  geom_col(aes(x = Pop, y = Old_New, fill = Old_New))+
  facet_wrap(~State, scales = "free", axis.labels = "all_x", ncol = 3)+
  scale_x_continuous(labels = scales::label_number(scale = .000001, suffix = "M"))+
  labs(y = "", x = "Population (Millions)", fill = "Class Change")
```




```{r urbanReclass, eval = FALSE}


# Save as centroids
# r2.sf <- cntrds%>%
#   left_join(pop, by = "GISJOIN")%>%
#   left_join(reclass.2.stats, by = c("GEOID20"="GEOID_Blk"))%>%
#   drop_na(NewClass)
# 
# st_write(r2.sf,here("Method_2/Data/reclass_urban.gpkg"))

table(reclass.2.stats$Pred_Class)
```


```{r tbl2, echo = FALSE}
reclass.2 <- vroom(here("Method_2/data/Reclass_2.csv"))

reclass.2.stats <- reclass.2%>%
  filter(NewClass == "Public" & is.na(HU_2020))

table(reclass.2.stats$Pred_Class)
```


Using this step, we reclassify 23,283 blocks from private to public.





# Combine New Dataset

```{r combine, eval = FALSE}
reclass.1.sel <- reclass.1%>%
  select(GEOID20,OldClass,NewClass,Old_New)%>%
  mutate(cat = "Complete State")%>%
  distinct()%>%
  filter(!is.na(NewClass))

reclass.2.sel <- reclass.2%>%
  select(GEOID_Blk,OldClass,NewClass,Old_New)%>%
  mutate(cat = "Urban Areas")%>%
  filter(!GEOID_Blk %in% reclass.1.sel$GEOID20)%>%
  distinct()%>%
  filter(!is.na(NewClass))
colnames(reclass.2.sel)[1] <- "GEOID20"

# reclass.3.sel <- reclass.3%>%
#   select(GEOID_Blk,OldClass,NewClass,Old_New)%>%
#   mutate(cat = "Partial / No State")%>%
#   filter(!GEOID_Blk %in% reclass.1.sel$GEOID20)%>%
#   filter(!GEOID_Blk %in% reclass.2.sel$GEOID20)%>%
#   distinct()
# colnames(reclass.3.sel)[1] <- "GEOID20"
# 
# reclass.all <- reclass.1.sel%>%
#   rbind(reclass.2.sel,reclass.3.sel)

reclass.combo <- rbind(reclass.1.sel,reclass.2.sel)
reclass.all <- ws.df%>%
  left_join(reclass.combo, by = c("GEOID_Blk"="GEOID20"))%>%
  mutate(Reclass_Type = ifelse(cat == "Complete State" & !OldClass == NewClass,"Complete State",
                          ifelse(cat == "Urban Area" & !OldClass == NewClass,"Urban Area - No Pop",
                                 ifelse(wasIsland == TRUE,"Was Island",
                                        ifelse(NA_Reclass == TRUE,"Was NA","None")))),
         Reclass_Type = replace_na(Reclass_Type,"None"),
         Final_Class = ifelse(!is.na(NewClass),NewClass,Pred_Class))%>%
  select(GEOID_Blk,GEOID_BlkGrp,State,HU_2020,Prob_Pub,Tree_Cat,Final_Class,Reclass_Type)

# Save File
vroom_write(reclass.all,here("outputs/Final_Classification.csv"), delim = ",")


# Save spatial file of centroids to inspect
reclass.sf <- vroom("D:/Github/ORD_Water_Supply/data/blocks_2020_all_centroids.csv")%>%
  left_join(pop, by = "GISJOIN")%>%
  left_join(ws.df, by = c("GEOID20"="GEOID_Blk"))%>%
  left_join(reclass.all, by = "GEOID20")%>%
  st_as_sf(coords = c("X","Y"), crs = st_crs(4326))

st_write(reclass.sf, here("Method_2/data/Method_2.gpkg"), layer = "Centroids_Reclass", append = FALSE)

reclass.all.info <- reclass.sf%>%
  st_drop_geometry()
vroom_write(reclass.all.info, here("Method_2/data/Reclass_all_info.csv"), delim = ",", append = FALSE)
```

## Reclassification Results

```{r reclassPlot}
reclass.all.info <- vroom(here("Method_2/data/Reclass_all_info.csv"))

reclass.stats <- reclass.all.info%>%
  group_by(Old_New)%>%
  summarise(n = n())

ggplot(reclass.stats)+
  geom_col(aes(x = Old_New, y = n/1000000, fill = Old_New), color = "black")+
  labs(title = "Reclassification", x = "Reclassification Result", y = "# Blocks (Millions)")+
  theme(legend.position = "none")
```

## Compare Well / Public Water Estimates
This is broken down into two categories:

- Original Estimate
- Binary Estimate With Reclassification

```{r}
counts <- reclass.all.info%>%
  mutate(B_Wells = ifelse(NewClass == "Private",HU_2020,0),
         B_Public = ifelse(NewClass == "Public",HU_2020,0))%>%
  summarise(O_Wells = sum(Est_Wells_2020,na.rm = TRUE),
            O_Public = sum(HU_2020-Est_Wells_2020,na.rm=TRUE),
            B_Wells = sum(B_Wells, na.rm=TRUE),
            B_Public = sum(B_Public,na.rm = TRUE))%>%
  pivot_longer(everything(),names_to = "var", values_to = "val")

counts$group = c("Wells","Public","Wells","Public")
counts$label <- c("Original","Original","Reclass","Reclass")
ggplot(counts)+
  geom_col(aes(x = label, y = val/1000000, group = group))+
  facet_wrap(~group, scales = "free")+
  labs(title="Comparison of Estimated Totals", x = "", y = "Housing Units (Millions)")
```

Prior to this reclassification, we estimated that `r round(100*(115009937/(115009937+25487797)),1)`% of housing units were using public water. After reclassification, we estimate that `r round(100*(120221224/(120221224+20276510)),1)`


```{r}
final <- reclass.all.info%>%
  mutate(NewClass = ifelse(Pred_Class == "Private" & wasIsland == TRUE & NewClass == "Private","Public",NewClass))

private <- final%>%
  filter(NewClass == "Private")

public <- final%>%
  filter(NewClass == "Public")

check <- final%>%
  filter(wasIsland == TRUE & Pred_Class == "Private" & Old_New == "Private -> Private")


wi <- final%>%
  filter(substr(GISJOIN,2,3)=="55")

wi.private <- wi%>%
  filter(NewClass == "Private")

wi.public <- wi%>%
  filter(NewClass == "Public")

print(paste0("Housing Units Self-Supplying Water: ",sum(private$U7G001,na.rm=TRUE)))
print(paste0("Housing Units on Public Supply: ", sum(public$U7G001,na.rm=TRUE)))
```

## 1990 data check
```{r}
hu.100 <- vroom("D:/data/nhgis/tables/Block_Groups/nhgis0293_ds120_1990_blck_grp.csv")%>%
  select(GISJOIN,ET1001,ESA001)%>%
  setNames(c("GISJOIN","Population","Housing_Units"))

sow <- vroom("D:/data/nhgis/tables/Block_Groups/nhgis0292_ds123_1990_blck_grp.csv")%>%
  select(GISJOIN,EX5001,EX5002,EX5003,EX5004)%>%
  mutate(pctselfSupply = (EX5002+EX5003+EX5004)/(EX5001+EX5002+EX5003+EX5004))

use90 <- hu.100%>%
  left_join(sow)%>%
  mutate(Pop_Self = Population*pctselfSupply)

use90.wi <- use90%>%
  filter(substr(GISJOIN,2,3)=="55")

# Import data for 1990 Census Block Groups
bg.90 <- vroom(here("data/Census_Tables/US_Blk_Grps_1990.csv"))%>%
  left_join(hu.100,by = c("GISJOIN_BG"="GISJOIN"))
```

